---
title: "GeoMX_placenta_analysis"
author: "Jose J Moreno-Villena"
date: "2025-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Placenta Spatial Transcriptome analysis using GeoMx Digital Spatial Profiler 

##GeoMX experiment at Queens Mary University London. The GeoMX NGS pipeline was run by the QMUL facility. This analysis begin with DCC files (digital count conversion files)

###Followed the vignette: 
####https://bioconductor.org/packages/release/workflows/vignettes/GeoMxWorkflows/inst/doc/GeomxTools_RNA-NGS_Analysis.html


##### install and load the GeoMx packages
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# The following initializes most up to date version of Bioc
BiocManager::install()

BiocManager::install("NanoStringNCTools")
BiocManager::install("GeomxTools")
BiocManager::install("GeoMxWorkflows")

library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)
```


Loading Data

```{r}
#the DataLocation folder should contain a dccs, pkcs, and annotation
mydatadir <- file.path("~/ucl/science/current_projects/Spatial_comparison/qmul_analysis")

# automatically list files in each directory for use
myDCCFiles <- dir(file.path(mydatadir,"DCC-20230921"), pattern = ".dcc$",
                full.names = TRUE, recursive = TRUE)

# probe assay metadata describing the gene targets present in the data
myPKCFiles <- paste0(mydatadir, "/Hsa_WTA_v1.0.pkc")

# Sample metadata (has to be an excel file) 
mySampleAnnotationFile <- file.path(mydatadir,"GC-TX-10347_sample_attributes_v2.xlsx")

# load data
myData <-
  readNanoStringGeoMxSet(dccFiles = myDCCFiles,
                         pkcFiles = myPKCFiles,
                         phenoDataFile = mySampleAnnotationFile,
                         phenoDataSheet = "Template",
                         phenoDataDccColName = "Sample_ID",
                         protocolDataColNames = c("aoi", "roi"),
                         experimentDataColNames = c("panel"))


```

Sample Overview: Sankey plot

```{r}
#####
library(dplyr)
library(ggforce)
library(networkD3)

sankeyCols <- c("source", "target", "value")

link1 <- count(pData(myData), GA,Condition)
link2 <- count(pData(myData),  Condition, ROIsampleID)
link3 <- count(pData(myData),  ROIsampleID, ROISubsection)

colnames(link1) <- sankeyCols
colnames(link2) <- sankeyCols
colnames(link3) <- sankeyCols

links <- rbind(link1,link2,link3)
nodes <- unique(data.frame(name=c(links$source, links$target)))

# sankeyNetwork is 0 based, not 1 based
links$source <- as.integer(match(links$source,nodes$name)-1)
links$target <- as.integer(match(links$target,nodes$name)-1)


sankeyNetwork(Links = links, Nodes = nodes, Source = "source",
              Target = "target", Value = "value", NodeID = "name",
              units = "TWh", fontSize = 12, nodeWidth = 30)

```

QC & Pre-processing

```{r}
# Shift counts to one
myData <- shiftCountsOne(myData, useDALogic = TRUE)


# Default QC cutoffs are commented in () adjacent to the respective parameters
# study-specific values were selected after visualizing the QC results in more
# detail below
QC_params <-
  list(minSegmentReads = 1000, # Minimum number of reads (1000)
       percentTrimmed = 80,    # Minimum % of reads trimmed (80%)
       percentStitched = 80,   # Minimum % of reads stitched (80%)
       percentAligned = 75,    # Minimum % of reads aligned (80%)
       percentSaturation = 50, # Minimum sequencing saturation (50%)
       minNegativeCount = 1,   # Minimum negative control counts (10)
       maxNTCCount = 9000,     # Maximum counts observed in NTC well (1000)
       minNuclei = 20,         # Minimum # of nuclei estimated (100)
       minArea = 1000)         # Minimum segment area (5000)

myData <-
  setSegmentQCFlags(myData, 
                    qcCutoffs = QC_params)   


# Collate QC Results
QCResults <- protocolData(myData)[["QCFlags"]]
flag_columns <- colnames(QCResults)
QC_Summary <- data.frame(Pass = colSums(!QCResults[, flag_columns]),
                         Warning = colSums(QCResults[, flag_columns]))
QCResults$QCStatus <- apply(QCResults, 1L, function(x) {
  ifelse(sum(x) == 0L, "PASS", "WARNING")
})
QC_Summary["TOTAL FLAGS", ] <-
  c(sum(QCResults[, "QCStatus"] == "PASS"),
    sum(QCResults[, "QCStatus"] == "WARNING"))

# Plot all of the QC Summary information in a table.
kable(QC_Summary, caption = "QC Summary Table for each Segment")


```

Remove flagged segments

```{r}

myData <- myData[, QCResults$QCStatus == "PASS"]

# Subsetting our dataset has removed samples which did not pass QC
dim(myData)

```

Probe QC: remove low-performing probes
A probe is removed globally from the dataset if either of the following is true:

    the geometric mean of that probe’s counts from all segments divided by the geometric mean of all probe counts representing the target from all segments is less than 0.1
    the probe is an outlier according to the Grubb’s test in at least 20% of the segments

A probe is removed locally (from a given segment) if the probe is an outlier according to the Grubb’s test in that segment.
```{r}

# Generally keep the qcCutoffs parameters unchanged. Set removeLocalOutliers to 
# FALSE if you do not want to remove local outliers
myData <- setBioProbeQCFlags(myData, 
                               qcCutoffs = list(minProbeRatio = 0.1,
                                                percentFailGrubbs = 20), 
                               removeLocalOutliers = TRUE)

ProbeQCResults <- fData(myData)[["QCFlags"]]

# Define QC table for Probe QC
qc_df <- data.frame(Passed = sum(rowSums(ProbeQCResults[, -1]) == 0),
                    Global = sum(ProbeQCResults$GlobalGrubbsOutlier),
                    Local = sum(rowSums(ProbeQCResults[, -2:-1]) > 0
                                & !ProbeQCResults$GlobalGrubbsOutlier))

qc_df 

#Subset object to exclude all that did not pass Ratio & Global testing
ProbeQCPassed <- 
  subset(myData, 
         fData(myData)[["QCFlags"]][,c("LowProbeRatio")] == FALSE &
           fData(myData)[["QCFlags"]][,c("GlobalGrubbsOutlier")] == FALSE)
dim(ProbeQCPassed)
#> Features  Samples 
#>    18641      229
myData <- ProbeQCPassed 

```

Create Gene-level Count Data

```{r}
# Check how many unique targets the object has
length(unique(featureData(myData)[["TargetName"]]))
#> [1] 18504

# collapse to targets
target_myData <- aggregateCounts(myData)
dim(target_myData)

exprs(target_myData)[1:5, 1:2]
```

Limit of Quantification & Filtering: determine the limit of quantification (LOQ) per segment. The LOQ is calculated based on the distribution of negative control probes and is intended to approximate the quantifiable limit of gene expression per segment. 


```{r}
#####

# Define LOQ SD threshold and minimum value
cutoff <- 2
minLOQ <- 2

# Calculate LOQ per module tested
LOQ <- data.frame(row.names = colnames(target_myData))
for(module in modules) {
  vars <- paste0(c("NegGeoMean_", "NegGeoSD_"),
                 module)
  if(all(vars[1:2] %in% colnames(pData(target_myData)))) {
    LOQ[, module] <-
      pmax(minLOQ,
           pData(target_myData)[, vars[1]] * 
             pData(target_myData)[, vars[2]] ^ cutoff)
  }
}
pData(target_myData)$LOQ <- LOQ

# Filtering: 
LOQ_Mat <- c()
for(module in modules) {
  ind <- fData(target_myData)$Module == module
  Mat_i <- t(esApply(target_myData[ind, ], MARGIN = 1,
                     FUN = function(x) {
                       x > LOQ[, module]
                     }))
  LOQ_Mat <- rbind(LOQ_Mat, Mat_i)
}
# ensure ordering since this is stored outside of the geomxSet
LOQ_Mat <- LOQ_Mat[fData(target_myData)$TargetName, ]



# Save detection rate information to pheno data
pData(target_myData)$GenesDetected <- 
  colSums(LOQ_Mat, na.rm = TRUE)
pData(target_myData)$GeneDetectionRate <-
  pData(target_myData)$GenesDetected / nrow(target_myData)

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
pData(target_myData)$DetectionThreshold <- 
  cut(pData(target_myData)$GeneDetectionRate,
      breaks = c(0, 0.01, 0.05, 0.1, 0.15, 1),
      labels = c("<1%", "1-5%", "5-10%", "10-15%", ">15%"))

# stacked bar plot of different cut points (1%, 5%, 10%, 15%)
ggplot(pData(target_myData),
       aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = ROISubsection)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
       y = "Segments, #",
       fill = "Segment Type")

```

Remove segments with less than 10% of the genes detected. 

```{r}
target_myData <-
  target_myData[, pData(target_myData)$GeneDetectionRate >= .1]

dim(target_myData)
```

None is removed


Gene Filtering

```{r}
# Subset to target genes detected in at least 10% of the samples.
#   Also manually include the negative control probe, for downstream use
negativeProbefData <- subset(fData(target_myData), CodeClass == "Negative")
neg_probes <- unique(negativeProbefData$TargetName)
target_myData <- 
  target_myData[fData(target_myData)$DetectionRate >= 0.1 |
                    fData(target_myData)$TargetName %in% neg_probes, ]
dim(target_myData)

```


Normalization
```{r}
###
# Q3 norm (75th percentile) for WTA/CTA  with or without custom spike-ins
target_myData <- normalize(target_myData ,
                             norm_method = "quant", 
                             desiredQuantile = .75,
                             toElt = "q_norm")

# Background normalization for WTA/CTA without custom spike-in
target_myData <- normalize(target_myData ,
                             norm_method = "neg", 
                             fromElt = "exprs",
                             toElt = "neg_norm")
```

